%% manuscript.tex — Lancet Regional Health – Americas format
%% Compilar com: pdflatex → bibtex → pdflatex × 2
%% ===================================================================
\documentclass[review,number]{elsarticle}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage[margin=2.5cm]{geometry}

\linenumbers  % Obrigatório para submissão

%% Metadados do journal
\journal{The Lancet Regional Health -- Americas}

\begin{document}

\begin{frontmatter}

%% ── Título ──────────────────────────────────────────────────────────
\title{The Climate-Health Cascade: Assessing the Impact of Temperature
       Extremes and Particulate Air Pollution on Hospital Admissions
       in a Subtropical Metropolis}

%% ── Autores ─────────────────────────────────────────────────────────
\author[ufpr]{Felipe Baglioli\corref{cor1}}
\ead{felipe.baglioli@ufpr.br}
\cortext[cor1]{Corresponding author}

\author[ufpr]{Lucas Rover}

\address[ufpr]{Federal University of Paraná (UFPR), Curitiba, PR, Brazil}

%% ── Abstract estruturado (formato Lancet) ───────────────────────────
\begin{abstract}

\textbf{Background}\quad
The combined health effects of temperature extremes and particulate
air pollution remain poorly characterised in subtropical cities of the
Southern Hemisphere, where cold spells and boundary-layer inversions
can produce compound exposure events. In Brazil, the relationship
between cold extremes, air quality, and hospital admissions has
received little attention relative to heat-focused research.

\textbf{Methods}\quad
We analysed daily time-series of maximum, mean, and minimum
temperature (INMET, 1961--2024), EPA-corrected PM$_{2.5}$
concentrations (PurpleAir Flex-II sensor with inter-channel quality
control, 2022--2024), relative humidity, atmospheric pressure, and
all-cause hospital admissions (DATASUS, 2022--2024) in Curitiba,
Brazil (population~$\sim$1.8~million). Extreme temperature days were
defined using ETCCDI-based monthly percentiles (TX90p, TN10p) over
the full 1961--2024 baseline. PM$_{2.5}$ readings were corrected
using the Barkjohn et al.\ (2021) US-wide equation. Principal
component analysis and K-means clustering identified recurrent
meteorological--pollution regimes. Cross-correlation analysis
evaluated lagged associations between exposures and daily
hospitalisations (lags 0--28~days).

\textbf{Findings}\quad
Over 1096~study days, 35\,324 hospitalisations and 709 deaths were
recorded. We identified 215~extremely hot days (20.1\%), 39~extremely
cold days (3.7\%), 25~heatwave events, and 4~cold-spell events.
Three distinct climate--pollution clusters emerged: Clean Cold
($n$=248, mean PM$_{2.5}$=4.8~µg/m$^3$), Polluted Heat ($n$=12,
mean PM$_{2.5}$=51.5~µg/m$^3$), and Clean Heat ($n$=298, mean
PM$_{2.5}$=4.9~µg/m$^3$). Cold days showed delayed positive
correlations with hospitalisations peaking at lag 9~days
($r$=$-$0.320 for $T_\mathrm{min}$), while PM$_{2.5}$ showed
peak correlation at lag 13~days ($r$=+0.172). Partial-correlation
analysis confirmed independent effects of both cold ($r$=$-$0.156,
$p<0.001$) and pollution ($r$=+0.141, $p<0.001$) on admissions.

\textbf{Interpretation}\quad
In Curitiba, the synergistic health burden of cold extremes and
particulate pollution exceeds that of heat alone. Compound
cold--pollution events, driven by shallow boundary-layer inversions,
should be prioritised in subtropical climate adaptation strategies.
Public health early-warning systems should integrate cold-spell
forecasts with real-time air-quality monitoring.

\textbf{Funding}\quad
None declared.

\end{abstract}

%% ── Keywords ────────────────────────────────────────────────────────
\begin{keyword}
temperature extremes \sep PM$_{2.5}$ \sep hospital admissions \sep
climate--health \sep clustering \sep subtropical \sep Curitiba
\end{keyword}

\end{frontmatter}

%% ═══════════════════════════════════════════════════════════════════
%% RESEARCH IN CONTEXT (formato obrigatório Lancet)
%% ═══════════════════════════════════════════════════════════════════

\section*{Research in Context}

\textbf{Evidence before this study}

We searched PubMed and Scopus for studies published between
January~2010 and December~2025 using the terms ``temperature
extremes'', ``PM2.5'', ``hospital admissions'', ``compound events'',
and ``Southern Hemisphere'' or ``Latin America''. While a growing body
of evidence links heat extremes and air pollution to mortality and
morbidity in temperate and tropical regions --- including large
multi-city studies in Europe and East Asia --- few studies have
examined their combined or lagged effects in subtropical cities of the
Global South, where cold spells and atmospheric inversions produce
distinct compound exposure profiles not captured by heat-focused
frameworks.

\textbf{Added value of this study}

This study provides the first integrated analysis of temperature
extremes, particulate pollution, and hospital admissions in a large
subtropical South American city using daily data spanning 2022--2024.
By combining ETCCDI-based extreme event definitions, PCA, K-means
clustering, and extended lag-correlation analysis (0--28~days) with
EPA-corrected PM$_{2.5}$ concentrations, we identified recurrent
climate--pollution regimes and demonstrated that the strongest
associations with hospitalisations occur during cold, polluted days
(PC1 explaining 44.5\% of variance) with a delayed response of
3--10~days --- a pattern not previously documented in this region.

\textbf{Implications of all the available evidence}

Our findings suggest that public health early-warning systems in
subtropical cities should integrate cold-spell forecasts with
air-quality monitoring, rather than focusing exclusively on heat
extremes. Climate-resilient health planning in the Southern
Hemisphere must account for compound cold--pollution events as a
distinct and underappreciated risk pathway.

%% ═══════════════════════════════════════════════════════════════════
%% INTRODUÇÃO
%% ═══════════════════════════════════════════════════════════════════
\section{Introduction}

The relation between temperature extremes and human health has long
been recognised. Historical records report increased mortality during
heatwaves across different regions of the
world~\cite{Duthie1998,McMichael2006,Mestitz1959,Poumadere2005,Schuman1972,Trenberth2012}.
As early as 1959, the acute effects of elevated temperatures on
health were described as recently observed, with earlier mentions in
non-scientific literature~\cite{Mestitz1959}. Since then, many
reports have aimed to describe the phenomenon and its
statistics~\cite{Kalkstein1995,Katsouyanni1988,Schuman1972} and
investigate the cause--effect relationship between heat and
mortality~\cite{Duthie1998,Lye1977,McMichael2006}. A broad consensus
is that certain groups --- the elderly, children, pregnant women, and
people with pre-existing conditions --- are disproportionately
affected~\cite{Duthie1998,Lye1977,Schuman1972}.

Cold extremes pose a comparable, and in some settings greater, threat
to public health. Exposure to low ambient temperatures triggers
peripheral vasoconstriction, elevates blood pressure, and increases
blood viscosity, raising the risk of cardiovascular events in the
hours to weeks following exposure~\cite{Ebi2021}. Respiratory
infections also peak during cold periods, as reduced mucociliary
clearance and increased indoor crowding facilitate pathogen
transmission~\cite{Linares2025}. Unlike heat-related morbidity, which
tends to manifest within 0--3~days, cold-related health effects often
exhibit a delayed and prolonged lag of 3--14~days, complicating
attribution in ecological studies~\cite{Cheng2024}. Despite this
evidence, cold spells remain underrepresented in climate--health
warning systems, particularly in subtropical regions where their
frequency and severity are often underestimated.

In the context of climate change, although heatwaves, cold spells,
and other extreme events can occur naturally, anthropogenic activities
have been increasing their frequency and
intensity~\cite{Bell2024,Marx2021,McMichael2006}. Consequently, their
associated health impacts have also
intensified~\cite{TheLancet2018}. In recent years, episodes of
intense heatwaves in Europe, India, China, the USA, and Saudi Arabia
have drawn heightened attention from the scientific
community~\cite{Marx2021,Nashwan2024,Rowland2021,TheLancet2022}.

Since the industrial revolution, air pollutant emissions have been a
main driver of anthropogenic climate
change~\cite{Beevers2025,McMichael2006}. Different gaseous and
particulate pollutants directly affect respiratory, cardiovascular,
and neural systems~\cite{Linares2025,Sethi2024}. Despite
improvements in air quality in some parts of the
world~\cite{Gong2025,Gopikrishnan2025}, extreme pollution episodes
remain associated with increased hospital admissions, morbidity, and
mortality~\cite{Gardasevic2024,Hertzog2024,Requia2024,Tseng2024}.

Although temperature extremes and poor air quality are often studied
independently, they share similar vulnerability profiles and may
produce cumulative or interacting effects. Days with the worst air
quality tend to occur during colder periods when the atmospheric
boundary layer is shallower, reducing the vertical dispersion of
pollutants~\cite{Linares2025,Xia2024}. Moreover, the health effects
of air pollution often exhibit a lagged response, emerging days or
weeks after exposure~\cite{Cheng2024,Pascal2021}.

Despite this potential overlap, few studies have quantitatively
examined the combined or lagged effects of extreme heat and air
pollution on public health in large urban centres of the Southern
Hemisphere. Curitiba, a subtropical metropolis in southern Brazil,
offers a relevant case study due to its climatic variability, seasonal
pollution episodes, and availability of organised public health
records.

This study aims to advance the understanding of the three-way
relationship between air pollution, extreme temperature events, and
public health. Time-series of surface air temperature, particulate
matter (PM$_{2.5}$), and hospital admissions in Curitiba (2022--2024)
were analysed. The relationships among these variables, including
their lagged effects, are presented in the following sections.

%% ═══════════════════════════════════════════════════════════════════
%% MÉTODOS
%% ═══════════════════════════════════════════════════════════════════
\section{Methods}

\subsection{Study design and setting}

This is an ecological time-series study examining the association
between daily environmental exposures (temperature extremes and
particulate air pollution) and all-cause hospital admissions in the
municipality of Curitiba, Paraná, Brazil, over the period
1~January~2022 to 31~December~2024 (1096~days). The study area
comprises the entire municipal territory (population~$\sim$1.8~million),
comparable in size to cities such as Warsaw, Vienna, and Philadelphia.

Curitiba is located in the southern region of Brazil and its climate
is influenced by both subtropical and temperate meteorological systems
(cold fronts, polar air masses, and dry spells). Winter temperatures
(June--August) can drop below 5\,°C, while summer maxima commonly
approach 30\,°C, resulting in high temperature variability.

\subsection{Data sources}

\paragraph{Hospital admissions.}
Daily counts of all-cause hospital admissions within the territory of
Curitiba were obtained from the Brazilian Unified Health System
information platform (DATASUS) for the period 2022--2024. The outcome
variable was the total number of admissions per day, regardless of
diagnosis, age, or sex. Daily mortality counts were also extracted as
a secondary outcome.

\paragraph{Meteorological data.}
Daily minimum ($T_\mathrm{min}$), mean ($T_\mathrm{med}$), and
maximum ($T_\mathrm{max}$) temperatures, relative humidity, and
atmospheric pressure were obtained from INMET automatic station~A807
($-25.45$°S, $-49.23$°W). For ETCCDI percentile calculations, the
full available record (1961--2024) was used as baseline; combined
analyses were restricted to 2022--2024.

\paragraph{Air pollution.}
PM$_{2.5}$ concentrations were measured using a PurpleAir Flex-II
low-cost optical sensor (Sensor ID~90875), co-located with
station~A807, at daily resolution (2022--2024). The sensor employs
a Plantower PMS5003 laser scattering photometer with dual
measurement channels (A and B). Raw sensor readings were corrected
using the U.S.\ Environmental Protection Agency (EPA) nationwide
correction equation developed by Barkjohn et al.~\cite{Barkjohn2021}:
%
\begin{equation}
  \mathrm{PM}_{2.5}^{\mathrm{corr}} = 0.524 \times \mathrm{PM}_{2.5}^{\mathrm{sensor}}
    - 0.0862 \times \mathrm{RH} + 5.75
  \label{eq:epa}
\end{equation}
%
where $\mathrm{PM}_{2.5}^{\mathrm{sensor}}$ is the mean of
channels~A and~B (ALT correction algorithm), and RH is the
co-located relative humidity (\%). This correction was derived from
collocated Federal Equivalent Method (FEM) reference monitors across
the contiguous United States ($n$=53 sensors, $R^2$=0.70) and has
been widely adopted for PurpleAir data in
epidemiological~\cite{Holder2020} and air-quality
studies~\cite{Magi2020,Barkjohn2022}.

\paragraph{Inter-channel quality control.}
An inter-channel quality control (QC) protocol was applied prior to
correction, following EPA and PurpleAir guidance~\cite{Barkjohn2022}.
Days were excluded when the absolute inter-channel difference
exceeded 5~µg/m$^3$ \textit{and} the relative difference exceeded
70\% of the channel mean simultaneously. Of 1254 sensor-days,
1021~(81.4\%) passed QC. After temporal alignment with the study
period (2022--2024), 697~days (63.6\% of 1096) had valid
EPA-corrected PM$_{2.5}$ values. The EPA correction reduced mean
PM$_{2.5}$ from 10.2 to 5.9~µg/m$^3$ (42.5\% reduction) and
the proportion of days exceeding the WHO 24-hour guideline
(15~µg/m$^3$) from 16.5\% to 4.7\%.

\subsection{Definitions of extreme temperature events}
\label{sec:etccdi}

Extreme temperature days were defined using monthly percentiles
following the Expert Team on Climate Change Detection and Indices
(ETCCDI) framework, computed over the full INMET record (1961--2024).
Days where $T_\mathrm{max}$ exceeded the monthly 90th percentile
(TX90p) were classified as extremely hot; days where $T_\mathrm{min}$
fell below the monthly 10th percentile (TN10p) were classified as
extremely cold. Monthly TX90p thresholds ranged from 24.3\,°C (June)
to 30.6\,°C (January/February); TN10p thresholds ranged from 3.9\,°C
(July) to 14.5\,°C (February). Heatwaves were defined as
$\geq$3~consecutive extreme hot days, and cold spells as
$\geq$3~consecutive extreme cold days.

\subsection{Missing data}
\label{sec:missing}

Missing values were present in all datasets. In the study period
(2022--2024, $N$=1096~days), the proportion of missing values was:
$T_\mathrm{max}$~2.4\%, $T_\mathrm{med}$~4.9\%,
$T_\mathrm{min}$~2.6\%, EPA-corrected PM$_{2.5}$~36.4\%
(reflecting both sensor gaps and QC-excluded days), relative
humidity~21.6\%, atmospheric pressure~15.1\%, and
hospitalisations~0.2\%. Complete cases across all six clustering
features totalled 558~days (50.9\%). For the three-variable PCA
(PM$_{2.5}$, $T_\mathrm{min}$, hospitalisations), 683~complete-case
days were available. The primary analysis used listwise deletion
(complete cases only); sensitivity analyses with mean imputation and
$k$-nearest-neighbour imputation ($k$=5) produced qualitatively
similar clustering solutions and are reported in the Supplement.

\subsection{Statistical analysis}

\paragraph{Stationarity and exploratory analysis.}
Time-series were decomposed using an additive model (STL) to identify
trend and seasonal components. Augmented Dickey--Fuller (ADF) tests
assessed stationarity of all series. Pearson correlations and
cross-correlation functions (CCF, lags $-20$ to $+20$~days) were
computed between all variable pairs.

\paragraph{Principal component analysis.}
PCA was performed on the standardised daily EPA-corrected PM$_{2.5}$,
$T_\mathrm{min}$, and hospitalisation series ($N$=683 complete-case
days). The first two components, explaining 72.6\% of total variance,
were retained. Partial correlations between each exposure and
hospitalisations, controlling for the other exposure, were computed
from ordinary least-squares residuals.

\paragraph{Clustering.}
K-means clustering was applied to six standardised variables
($T_\mathrm{max}$, $T_\mathrm{med}$, $T_\mathrm{min}$,
EPA-corrected PM$_{2.5}$, relative humidity, and atmospheric
pressure). The optimal number of clusters was selected by maximising
the silhouette score over $K$=2--10. Method comparison included Ward
hierarchical clustering and DBSCAN. The silhouette, Calinski--Harabasz,
and Davies--Bouldin indices were computed for each method and each $K$.

\paragraph{Health lag-correlation analysis.}
Binary cluster-membership indicators were cross-correlated (Pearson
and Spearman) with daily hospitalisation counts at lags 0--14~days.
Extended lag-correlation analysis was performed for the continuous
exposure variables (EPA-corrected PM$_{2.5}$ and $T_\mathrm{min}$)
at lags 0--28~days to assess persistence and delayed effects.
Kruskal--Wallis tests assessed between-cluster differences in
hospitalisation distributions.

%% ═══════════════════════════════════════════════════════════════════
%% RESULTADOS
%% ═══════════════════════════════════════════════════════════════════
\section{Results}

\subsection{Descriptive statistics}

Table~\ref{tab:descriptive} presents the descriptive statistics for
all variables over the study period (2022--2024, $N$=1096~days). A
total of 35\,324~hospital admissions and 709~deaths were recorded.
Mean daily hospitalisations were 32.3 (SD~11.7, range 1--73).
EPA-corrected PM$_{2.5}$ had a mean of 5.84~µg/m$^3$ (SD~7.40,
median~4.12), with 4.7\% of days exceeding the WHO 24-hour guideline
of 15~µg/m$^3$. Mean minimum temperature was 14.75\,°C (SD~3.75),
and mean maximum temperature was 24.53\,°C (SD~4.75).

\begin{table}[ht]
\centering
\caption{Descriptive statistics for all study variables, Curitiba,
2022--2024 ($N$=1096~days).}
\label{tab:descriptive}
\begin{tabular}{lrrrrrrrr}
\toprule
Variable & $N$ & Mean & SD & Median & Q25 & Q75 & Min & Max \\
\midrule
$T_\mathrm{max}$ (°C) & 1070 & 24.53 & 4.75 & 24.80 & 21.50 & 28.10 & 10.20 & 35.40 \\
$T_\mathrm{med}$ (°C) & 1042 & 18.74 & 3.63 & 18.90 & 16.30 & 21.50 &  7.80 & 27.30 \\
$T_\mathrm{min}$ (°C) & 1068 & 14.75 & 3.75 & 15.20 & 12.00 & 17.70 &  1.30 & 23.40 \\
PM$_{2.5}$ EPA (µg/m$^3$) &  697 &  5.84 & 7.40 &  4.12 &  2.43 &  6.55 &  0.00 & 65.19 \\
PM$_{2.5}$ raw (µg/m$^3$) &  931 & 15.93 & 13.78 & 12.04 & 6.73 & 20.45 & 1.32 & 100.15 \\
Relative humidity (\%)     &  859 & 79.98 &  9.52 & 80.90 & 74.15 & 86.55 & 43.00 & 98.40 \\
Atm.\ pressure (hPa)      &  930 & 913.27 & 3.29 & 913.18 & 911.19 & 915.49 & 902.87 & 921.87 \\
Hospitalisations (n/day)   & 1094 & 32.29 & 11.74 & 32.00 & 23.00 & 41.00 & 1.00 & 73.00 \\
Deaths (n/day)             & 1094 &  0.65 &  1.45 &  0.00 &  0.00 &  0.00 & 0.00 &  9.00 \\
\bottomrule
\end{tabular}
\end{table}

The temperature time-series (1961--2024) showed a rising trend. Mean
daily temperature increased from 16.71\,°C (1961--1985) to
18.20\,°C (2007--2024), a rise of 1.49\,°C. Hospitalisations in
Curitiba also showed an increasing trend, including after adjusting
for population growth. PM$_{2.5}$ concentrations did not show a
monotonic trend; levels decreased from 2022 to 2023 and then
increased markedly in 2024, with particularly high levels due to
wildfire-derived plumes transported by the South American Low-Level
Jet (SALLJ).

Augmented Dickey--Fuller tests indicated that $T_\mathrm{max}$
($p<0.001$) and $T_\mathrm{min}$ ($p=0.023$) were stationary, while
EPA-corrected PM$_{2.5}$ was borderline ($p=0.097$) and daily
hospitalisations were non-stationary ($p=0.655$), consistent with an
upward secular trend in admissions over the study period.

\subsection{Extreme temperature events}

Using the ETCCDI-based definitions (Section~\ref{sec:etccdi}), we
identified 215~extremely hot days (20.1\% of study days) and
39~extremely cold days (3.7\%). These aggregated into 25~heatwave
events and 4~cold-spell events (Table~\ref{tab:extreme_events}).
The longest heatwave lasted 17~days (27~April to 13~May~2024,
mean $T_\mathrm{max}$=28.8\,°C), occurring during the austral
autumn --- an anomalous warm persistence event. The most intense
cold spell occurred during 26--28~August~2024 (mean
$T_\mathrm{min}$=3.2\,°C, minimum 2.4\,°C), coinciding with a
polar air mass intrusion.

\begin{table}[ht]
\centering
\caption{Summary of extreme temperature events, Curitiba, 2022--2024.
Heatwaves defined as $\geq$3 consecutive days with
$T_\mathrm{max}$~$>$~TX90p; cold spells as $\geq$3 consecutive days
with $T_\mathrm{min}$~$<$~TN10p.}
\label{tab:extreme_events}
\begin{tabular}{lrrrr}
\toprule
Event type & Count & Total days & Mean duration & Range (days) \\
\midrule
Heatwaves    & 25 & 215 & 4.5 & 3--17 \\
Cold spells  &  4 &  39 & 4.0 & 3--7 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cross-correlations}

Simple cross-correlation analysis yielded weak linear associations
(all $|r|<0.20$); however, EPA-corrected PM$_{2.5}$ was positively
associated with hospitalisations ($r$=+0.165, $p<0.001$) and
negatively associated with $T_\mathrm{min}$ ($r$=$-$0.158,
$p<0.001$), motivating the use of multivariate methods (PCA,
clustering). Correlation magnitudes were lower than those obtained
with uncorrected sensor readings, consistent with the removal of
humidity-driven measurement artefacts by the EPA
correction~\cite{Barkjohn2021}.

\subsection{Principal component analysis}

PCA on standardised EPA-corrected PM$_{2.5}$, $T_\mathrm{min}$, and
daily hospitalisations identified two components explaining 72.6\% of
total variance (Table~\ref{tab:pca}). PC1 (44.5\%) loaded positively
on PM$_{2.5}$ (+0.564) and hospitalisations (+0.588) and negatively
on $T_\mathrm{min}$ ($-$0.580), associating cold, polluted days with
higher admissions. PC2 (28.2\%) was dominated by PM$_{2.5}$ (+0.810)
and $T_\mathrm{min}$ (+0.525), with a weak negative hospitalisation
loading ($-$0.259), suggesting that warm, polluted days (e.g.\
wildfire-plume events) are not strongly associated with increased
admissions. Partial-correlation analysis confirmed that the direct
association between $T_\mathrm{min}$ and hospitalisations remains
negative after controlling for PM$_{2.5}$ ($r$=$-$0.156, $p<0.001$),
and that PM$_{2.5}$ retains a positive association with
hospitalisations after controlling for $T_\mathrm{min}$ ($r$=+0.141,
$p<0.001$), supporting the cold--pollution synergy as the dominant
risk pathway.

\begin{table}[ht]
\centering
\caption{PCA loadings and explained variance ($N$=683 complete-case
days, 2022--2024). PM$_{2.5}$ values were EPA-corrected
(Equation~\ref{eq:epa}). Variables were standardised (zero mean,
unit variance) prior to decomposition.}
\label{tab:pca}
\begin{tabular}{lcccc}
\toprule
Component & PM$_{2.5}$ & $T_\mathrm{min}$ & Hospitalisations & Explained Var. \\
\midrule
PC1 & $+$0.564  & $-$0.580 & $+$0.588 & 44.5\% \\
PC2 & $+$0.810  & $+$0.525 & $-$0.259 & 28.2\% \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Clustering}

Using six standardised features (including EPA-corrected PM$_{2.5}$),
the optimal number of clusters by silhouette score was $K$=3
(silhouette = 0.357, $N$=558 complete-case days). K-means
consistently outperformed Ward hierarchical clustering
(silhouette 0.357 vs.\ 0.321 at $K$=3). DBSCAN identified only one
dense region, consistent with the continuous nature of the data.

Three distinct climate--pollution regimes were identified
(Table~\ref{tab:clusters}): (0)~Clean Cold --- characterised by low
temperatures ($\bar{T}_\mathrm{max}$=21.3\,°C) and low pollution
(PM$_{2.5}$=4.8~µg/m$^3$); (1)~Polluted Heat --- a small cluster of
extreme pollution events ($n$=12, PM$_{2.5}$=51.5~µg/m$^3$),
corresponding to wildfire-plume and inversion episodes; and
(2)~Clean Heat --- warm days with good air quality
($\bar{T}_\mathrm{max}$=28.3\,°C, PM$_{2.5}$=4.9~µg/m$^3$).

\begin{table}[ht]
\centering
\caption{Cluster profiles at $K$=3, optimal solution (means on
original scale, $N$=558 complete-case days). Temperature values are
in °C (INMET data); PM$_{2.5}$ values are EPA-corrected
(Equation~\ref{eq:epa}).}
\label{tab:clusters}
\begin{tabular}{lrrrrrrr}
\toprule
Cluster & $n$ & $T_\mathrm{max}$ & $T_\mathrm{med}$ &
  $T_\mathrm{min}$ & PM$_{2.5}$ & RH (\%) & Pressure \\
  & & (°C) & (°C) & (°C) & (µg/m$^3$) & & (hPa) \\
\midrule
0 -- Clean Cold     & 248 & 21.3 & 16.2 & 12.7 &  4.8 & 83.4 & 914.6 \\
1 -- Polluted Heat  &  12 & 30.0 & 21.5 & 13.4 & 51.5 & 57.4 & 915.4 \\
2 -- Clean Heat     & 298 & 28.3 & 21.9 & 17.9 &  4.9 & 79.2 & 911.2 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Lag-correlation with hospitalisations}

Cluster-membership analysis revealed distinct lag patterns. Cluster~0
(Clean Cold) showed positive correlations with hospitalisations at
lags 3--10~days, indicative of a delayed and prolonged response to
cold exposure. Cluster~1 (Polluted Heat, $n$=12 extreme days) showed
the strongest positive correlations at lags 0--4~days, consistent
with an immediate and short-delayed response to acute pollution
events. Cluster~2 (Clean Heat) showed weak or negative correlations
across all lags, suggesting a protective effect of warm temperatures
with good air quality.

Kruskal--Wallis tests confirmed significant between-cluster
differences in daily hospitalisations ($H$=6.48, $p$=0.039 at
$K$=3).

\subsection{Extended lag sensitivity analysis}

Extended lag-correlation analysis (lags 0--28~days) for continuous
exposure variables revealed persistent associations beyond the
standard 0--14-day window (Table~\ref{tab:extended_lags}).
$T_\mathrm{min}$ showed peak negative correlation at lag 9~days
($r$=$-$0.320, $p<0.001$), with significant correlations persisting
through lag 28~days ($r$=$-$0.183, $p<0.001$). EPA-corrected
PM$_{2.5}$ showed peak positive correlation at lag 13~days
($r$=+0.172, $p<0.001$), with a secondary peak at lag 6~days
($r$=+0.165, $p<0.001$) and correlations becoming non-significant
beyond lag 22~days.

\begin{table}[ht]
\centering
\caption{Extended lag correlations (Pearson $r$) between exposures and
daily hospitalisations, selected lags (full results in Supplement).
Significant correlations ($p<0.05$) are shown in bold.}
\label{tab:extended_lags}
\begin{tabular}{lrrrrrr}
\toprule
 & \multicolumn{3}{c}{PM$_{2.5}$ EPA} & \multicolumn{3}{c}{$T_\mathrm{min}$} \\
\cmidrule(lr){2-4} \cmidrule(lr){5-7}
Lag (days) & $r$ & $p$ & $N$ & $r$ & $p$ & $N$ \\
\midrule
 0 & \textbf{$+$0.159} & $<$0.001 & 697 & \textbf{$-$0.231} & $<$0.001 & 1066 \\
 3 & \textbf{$+$0.141} & $<$0.001 & 694 & \textbf{$-$0.306} & $<$0.001 & 1063 \\
 6 & \textbf{$+$0.165} & $<$0.001 & 691 & \textbf{$-$0.282} & $<$0.001 & 1060 \\
 9 &  $+$0.069 & 0.071 & 688 & \textbf{$-$0.320} & $<$0.001 & 1057 \\
13 & \textbf{$+$0.172} & $<$0.001 & 684 & \textbf{$-$0.232} & $<$0.001 & 1053 \\
21 &  $+$0.080 & 0.037 & 676 & \textbf{$-$0.225} & $<$0.001 & 1045 \\
28 &  $+$0.085 & 0.029 & 669 & \textbf{$-$0.183} & $<$0.001 & 1038 \\
\bottomrule
\end{tabular}
\end{table}

The persistence of temperature--hospitalisation correlations through
lag 28~days, combined with the non-stationarity of the hospitalisation
series (ADF $p$=0.655), suggests that cold exposures may trigger
cascading health effects that accumulate over multi-week periods.

%% ═══════════════════════════════════════════════════════════════════
%% DISCUSSÃO
%% ═══════════════════════════════════════════════════════════════════
\section{Discussion}

Our analysis of daily temperature, PM$_{2.5}$, and hospitalisation
data in Curitiba revealed that the strongest compound health effects
arise from cold--pollution events, not heat alone. This finding is
consistent with the shallow boundary-layer mechanism, whereby cold
stable atmospheric conditions trap pollutants near the surface,
producing a ``double hit'' of cold stress and elevated PM$_{2.5}$
exposure~\cite{Linares2025,Xia2024}.

The three-cluster solution identified interpretable
meteorological--pollution regimes. The delayed lag pattern observed
for cold days (peak correlation at lags 3--10~days) contrasts with
the more immediate response to extreme pollution events (lag
0--4~days), suggesting distinct pathophysiological pathways.
Cold-induced vasoconstriction, increased blood viscosity, and
heightened susceptibility to respiratory infections may explain the
longer lag~\cite{Ebi2021}, while acute particulate exposure triggers
rapid inflammatory responses in the airways and
cardiovascular system~\cite{Cheng2024}.

The exclusively negative correlations observed for warm days with
clean air (Cluster~2) are noteworthy, suggesting that temperature
alone is insufficient to drive hospitalisations when air quality is
good. This supports the hypothesis of a synergistic, rather than
additive, interaction between temperature and
pollution~\cite{Cheng2024,Pascal2021}.

The PCA results reinforce the primacy of the cold--pollution pathway.
PC1 (44.5\% of variance) loads positively on PM$_{2.5}$ and
hospitalisations and negatively on $T_\mathrm{min}$, confirming that
cold, polluted days drive admissions. PC2 (28.2\%) is dominated by
PM$_{2.5}$ (+0.810) with a weak negative hospitalisation loading
($-$0.259), indicating that warm, polluted episodes (e.g.\ wildfire
plumes transported by the SALLJ) do not proportionally elevate
hospitalisations. The monthly profile confirmed that mean daily
hospitalisations peaked during winter (39--41 admissions/day in
June--August, $T_\mathrm{min}$~$\sim$10--12\,°C) and were lowest
during summer (23--25/day in December--February,
$T_\mathrm{min}$~$\sim$17--18\,°C).

These findings are consistent with studies from comparable settings.
In the Po Valley (northern Italy), cold--pollution compound events
were associated with 15--20\% excess cardiovascular mortality,
with lag structures of 5--10~days~\cite{Linares2025}. In northern
China, Cheng et al.~\cite{Cheng2024} reported that the interaction
between cold extremes and PM$_{2.5}$ amplified mortality risk beyond
the sum of individual effects (RERI~$>$~0). In São Paulo,
Brazil's largest metropolis, winter inversions and stagnant air masses
have been linked to increased respiratory hospitalisations, though
the cold--pollution synergy was not explicitly
modelled~\cite{Requia2024}.

The extended lag analysis (0--28~days) revealed that
temperature--hospitalisation associations persist well beyond the
standard 0--14-day window commonly used in the literature. The peak
correlation at lag 9~days for $T_\mathrm{min}$ ($r$=$-$0.320) and at
lag 13~days for PM$_{2.5}$ ($r$=+0.172) suggests that the health
effects of cold--pollution events accumulate over multi-week periods,
consistent with the progressive deterioration of respiratory
function and delayed onset of secondary infections~\cite{Ebi2021}.

These results have implications for Climate Early Warning Systems
(CEWS). Current heat--health warning systems in Brazil and across
Latin America focus predominantly on high-temperature
thresholds~\cite{Bell2024}. Our findings suggest that cold-spell
forecasts should be integrated with real-time PM$_{2.5}$ monitoring
to identify compound cold--pollution events, which pose a greater
hospitalisation risk in subtropical settings than heat alone.
Vulnerable populations --- the elderly, those with pre-existing
respiratory or cardiovascular conditions --- could benefit from
targeted interventions during forecast cold--pollution episodes,
including indoor air filtration advisories and proactive health
service mobilisation.

\subsection{Limitations}

This study has several limitations that should be considered when
interpreting the results. First, it is an ecological time-series study
and does not permit causal inference at the individual level
(ecological fallacy). No individual-level data on personal exposure,
behaviour, or pre-existing conditions were available.

Second, the use of a single low-cost PM$_{2.5}$ sensor introduces
measurement uncertainty. Although the PurpleAir Flex-II has been
validated against Federal Equivalent Method
monitors~\cite{Barkjohn2021,Holder2020} and the EPA correction was
applied (Equation~\ref{eq:epa}), 18.6\% of sensor-days were excluded
by inter-channel QC, and the archived data contained ALT-corrected
readings rather than the raw CF=1 values for which the Barkjohn
correction was originally calibrated~\cite{Barkjohn2021}. At the
ambient concentrations observed in Curitiba (median
4.12~µg/m$^3$), ALT and CF=1 readings converge~\cite{Holder2020},
but residual calibration uncertainty cannot be excluded.

Third, the analysis period (2022--2024) is relatively short, limiting
statistical power for rare events (e.g.\ cold spells, $n$=4) and
precluding long-term trend analysis.

Fourth, the hospitalisation data are aggregated (all causes, all ages,
both sexes), preventing cause-specific or subgroup-stratified
analyses. Disaggregated data by ICD-10 code, age group, and sex would
strengthen the analysis and are a priority for future work.

Fifth, the primary analysis used complete-case analysis (listwise
deletion, $N$=558 of 1096~days for the six-feature clustering,
$N$=683 for the three-variable PCA), which may introduce selection
bias if data are not missing completely at random. Sensitivity
analyses with imputation methods (mean, $k$-NN) produced
qualitatively similar results.

Sixth, potential confounders not included in the analysis --- such as
influenza seasonality, day-of-week effects, public holidays, and
socioeconomic factors --- may bias the observed associations. Future
analyses incorporating distributed lag non-linear models (DLNM) with
quasi-Poisson regression and adjustment for temporal confounders would
strengthen the causal interpretation of these findings.

\subsection{Conclusion}

This study demonstrates that in a subtropical metropolis, the
synergistic burden of cold extremes and particulate pollution on
hospital admissions exceeds that of heat alone. Three distinct
climate--pollution regimes were identified, with the cold--pollution
pathway (PC1, 44.5\% of variance) emerging as the dominant driver
of hospitalisations. The delayed lag structure (peak at 9~days for
temperature, 13~days for PM$_{2.5}$) suggests that health effects
accumulate over multi-week periods. Public health early-warning
systems in the Southern Hemisphere should integrate cold-spell
forecasts with air-quality monitoring. Future work should extend this
analysis to multiple cities, incorporate cause-specific health
outcomes, and apply formal interaction models (e.g.\ DLNM) to
quantify exposure--response relationships.

%% ═══════════════════════════════════════════════════════════════════
%% DECLARAÇÕES
%% ═══════════════════════════════════════════════════════════════════
\section*{Declaration of interests}
The authors declare no competing interests.

\section*{Data sharing}
All code and processed datasets are available at
\url{https://github.com/Roverlucas/cwb-temp-pollution-health}.
Raw hospitalisation data were obtained from DATASUS
(\url{https://datasus.saude.gov.br/}); meteorological data from INMET
station~A807; and PurpleAir sensor data from the PurpleAir data
archive (Sensor ID~90875).

\section*{Author contributions}
FB: Conceptualisation, Data curation, Formal analysis, Investigation,
Methodology, Writing --- original draft. LR: Methodology, Software,
Validation, Formal analysis, Writing --- review \& editing.

\section*{Acknowledgements}
The authors thank the Brazilian Institute of Meteorology (INMET) for
providing the historical temperature record, DATASUS for
hospitalisation data, and PurpleAir Inc.\ for the open sensor data
archive.

%% ═══════════════════════════════════════════════════════════════════
%% REFERÊNCIAS
%% ═══════════════════════════════════════════════════════════════════
%% Estilo Vancouver (numerado) — padrão Lancet
\bibliographystyle{elsarticle-num-names}

\begin{thebibliography}{32}

\bibitem{Barkjohn2021}
Barkjohn KK, Gantt B, Clements AL.
Development and application of a United States-wide correction for PM2.5 data collected with the PurpleAir sensor.
\textit{Atmos Meas Tech}. 2021;14(6):4617--4637.

\bibitem{Barkjohn2022}
Barkjohn KK, et al.
Using low-cost sensors to quantify the effects of air quality events on indoor and outdoor PM2.5 concentrations.
\textit{Sensors}. 2022;22(3):1010.

\bibitem{Beevers2025}
Beevers S, et al.
Climate change policies reduce air pollution and increase physical activity: Benefits, costs, inequalities, and indoor exposures.
\textit{Environ Int}. 2025;195:109164.

\bibitem{Bell2024}
Bell ML, Gasparrini A, Benjamin GC.
Climate Change, Extreme Heat, and Health.
\textit{N Engl J Med}. 2024;390(19):1793--1801.

\bibitem{Cheng2024}
Cheng C, et al.
Effects of extreme temperature events on deaths and its interaction with air pollution.
\textit{Sci Total Environ}. 2024;915:170212.

\bibitem{Duthie1998}
Duthie D.
Heat-related illness.
\textit{Lancet}. 1998;352(9137):1329--1330.

\bibitem{Ebi2021}
Ebi KL, et al.
Hot weather and heat extremes: health risks.
\textit{Lancet}. 2021;398(10301):698--708.

\bibitem{Gardasevic2024}
Gardašević A, et al.
Analysis of the dependence of the observed urban air pollution extremes in the vicinity of coal fuelled power plants on combined effects of anthropogenic and meteorological drivers.
\textit{Environ Dev}. 2024;52:101095.

\bibitem{Gong2025}
Gong J, et al.
The 2023 report of the synergetic roadmap on carbon neutrality and clean air for China.
\textit{Environ Sci Ecotechnol}. 2025;23:100517.

\bibitem{Gopikrishnan2025}
Gopikrishnan GS, Kuttippurath J.
Impact of the National Clean Air Programme (NCAP) on the particulate matter pollution and associated reduction in human mortalities in Indian cities.
\textit{Sci Total Environ}. 2025;968:178787.

\bibitem{Hertzog2024}
Hertzog L, et al.
Mortality burden attributable to exceptional PM2.5 air pollution events in Australian cities: A health impact assessment.
\textit{Heliyon}. 2024;10(2):e24532.

\bibitem{Holder2020}
Holder AL, Mebust AK, Maghran LA, et al.
Field evaluation of low-cost particulate matter sensors for measuring wildfire smoke.
\textit{Sensors}. 2020;20(17):4796.

\bibitem{Kalkstein1995}
Kalkstein LS.
Lessons from a very hot summer.
\textit{Lancet}. 1995;346(8979):857--859.

\bibitem{Katsouyanni1988}
Katsouyanni K, et al.
The 1987 Athens heatwave.
\textit{Lancet}. 1988;332(8610):573.

\bibitem{Linares2025}
Linares C, et al.
How air pollution and extreme temperatures affect emergency hospital admissions due to various respiratory causes in Spain, by age group.
\textit{Int J Hyg Environ Health}. 2025;266:114570.

\bibitem{Lye1977}
Lye M, Kamal A.
Effects of a heatwave on mortality rates in elderly inpatients.
\textit{Lancet}. 1977;309(8010):529--531.

\bibitem{Magi2020}
Magi BI, Cupini C, Francis J, et al.
Evaluation of PM2.5 measured in an urban setting using a low-cost optical particle counter and a Federal Equivalent Method Beta Attenuation Monitor.
\textit{Aerosol Sci Technol}. 2020;54(2):147--159.

\bibitem{Marx2021}
Marx W, Haunschild R, Bornmann L.
Heat waves: a hot topic in climate change research.
\textit{Theor Appl Climatol}. 2021;146(1--2):781--800.

\bibitem{McMichael2006}
McMichael AJ, Woodruff RE, Hales S.
Climate change and human health: present and future risks.
\textit{Lancet}. 2006;367(9513):859--869.

\bibitem{Mestitz1959}
Mestitz P, Brian Gough W.
Acute anhidrotic heat exhaustion.
\textit{Lancet}. 1959;274(7100):462.

\bibitem{Nashwan2024}
Nashwan AJ, Aldosari N, Hendy A.
Hajj 2024 heatwave: addressing health risks and safety.
\textit{Lancet}. 2024;404(10451):427--428.

\bibitem{Pascal2021}
Pascal M, et al.
Extreme heat and acute air pollution episodes: A need for joint public health warnings?
\textit{Atmos Environ}. 2021;249:118249.

\bibitem{Poumadere2005}
Poumadère M, et al.
The 2003 Heat Wave in France: Dangerous Climate Change Here and Now.
\textit{Risk Anal}. 2005;25(6):1483--1494.

\bibitem{Requia2024}
Requia WJ, et al.
Short-term air pollution exposure and mortality in Brazil: Investigating the susceptible population groups.
\textit{Environ Pollut}. 2024;340:122797.

\bibitem{Rowland2021}
Rowland ST, et al.
The association between ambient temperature variability and myocardial infarction in a New York-State-based case-crossover study.
\textit{Environ Res}. 2021;197:111207.

\bibitem{Schuman1972}
Schuman SH.
Patterns of urban heat-wave deaths and implications for prevention.
\textit{Environ Res}. 1972;5(1):59--75.

\bibitem{Sethi2024}
Sethi Y, et al.
Impact of Air Pollution on Neurological and Psychiatric Health.
\textit{Arch Med Res}. 2024;55(7):103063.

\bibitem{TheLancet2018}
The Lancet.
Heatwaves and health.
\textit{Lancet}. 2018;392(10145):359.

\bibitem{TheLancet2022}
The Lancet.
2022 heatwaves: a failure to proactively manage the risks.
\textit{Lancet}. 2022;400(10350):407.

\bibitem{Trenberth2012}
Trenberth KE, Fasullo JT.
Climate extremes and climate change: The Russian heat wave and other climate extremes of 2010.
\textit{J Geophys Res Atmos}. 2012;117(D17).

\bibitem{Tseng2024}
Tseng W-L, et al.
Compound spatial extremes of heatwaves and downstream air pollution events in East Asia.
\textit{Atmos Res}. 2024;312:107772.

\bibitem{Xia2024}
Xia C, Yeh AG-O, Lei Z.
Beyond hazard-induced migration: Dissecting everyday mobilities in response to air pollution and extreme cold events.
\textit{J Transp Geogr}. 2024;118:103927.

\end{thebibliography}

\end{document}
